Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-12-19 threshold/draft.Rnw
JSS style update

%
%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'biclust' package installed in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.
%%
\documentclass[10pt]{article}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

@@

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
University of Rhode Island
\and
Michael Kane\\
Yale University
\and
Bryan Lewis\\
Paradigm4, Inc.\\
\and
Alex Poliakov\\
Paradigm4, Inc.
}

\date{}

\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle

\begin{abstract}

Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.

\end{abstract}

@@

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes\cite{genomics}.  Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications\cite{ml1, ml2} and are important to time series and
image processing applications\cite{timeseries, svd-similarity}.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-occurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

Most of the floating point operations occur in step 4 of Algorithm \ref{prune}.
Fortunately the $\ell$ evaluations in step 4 are independent of each other and
can easily be computed in parallel. See the reference R
implementation\cite{sup} for an example.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlbar}, Matlab\cite{irlbam}, and Python\cite{irlbap}.

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without any constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The R implementation\cite{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

A full/na\"ive thresholded correlation computation requires $O(n^2 m)$ flops.
Let $p$ be the selected IRLBA dimension. Then Algorithm \ref{prune} requires
$O(n p \ell)$ flops, where $\ell$ is the longest run of ordered entries in
$v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
including the flops required by IRLBA and final thresholding in step 7.  The
main computational savings of $\frac{nm}{lp}$ depends on the rate of decay in
the singular values of $A$, the subspace dimension $p$, and the desired
threshold value. We remark that this estimate of computational savings is
available already in step 3 of Algorithm \ref{prune} after only modest
computational effort. Users might use this cheap estimate to, for example,
alter their threshold in some cases.

@@

We evaluated the algorithm using two public gene expression and methylation
datasets.  Most tests were performed on a desktop PC with a single AMD
A10-7850K 3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered
RAM running the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R
version 3.1.2 using double-precision arithmetic and the OpenBLAS library
version 0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=1}. The last experiment includes a test run on a Linux
cluster as described in the text.

All examples use the reference R implementation, {\tt tcor()}, which can be
installed for the R language from the development GitHub repository using the
{\tt devtools} package with:

@@

The native R {\tt cor()} function is a general function that can compute
several different types of correlation matrices including Pearson, Spearman,
and Kendall and includes a number of options for dealing with missing values.
Because of its flexibility, the native {\tt cor()} function does not always
compute Pearson correlation matrices in the most computationally-efficient way.

@@

\subsection{Small EisenYeast example}

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} R package\cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as a $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(irlba)
source("tcor.R")  # XXX Replace later with tcor pacakge XXX XXX
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
threshold = 0.95

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The following example uses calculated beta values from the Illumina
Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
samples obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}.
See the supplemental code\cite{sup} for examples of efficiently converting the
downloaded sample methylation data into an R matrix. We chose these data for
their public availability as a general example of computing thresholded
correlation among sets of large numbers of vectors.

@@

@@

We illustrate that steps 4--7 of Algorithm \ref{prune} can be computed in
parallel by comparing the above results run on our reference test PC with
results run on a modest GNU/Linux cluster at Paradigm4, Inc. The cluster
consisted of four computers connected by 10 gigabit ethernet. Each computer
contained two Intel Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per
computer) and 64$\,$GB/1,600$\,$MHz ECC registered RAM running CentOS 6.6 and
64-bit R version 3.2.2 with the reference R BLAS library.

No code change was required to run in parallel on the cluster. The reference R
{\tt tcor()} implementation uses R's {\tt foreach}\cite{foreach} framework to
run on arbitrary parallel configurations (``back ends''). We simply registered
a {\tt doRedis}\cite{doredis} parallel back end with 16 R workers per computer
(64 total R worker processes) prior to invoking {\tt tcor()}.

The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
below. The cluster example computed the filtering step 7 of Algorithm
\ref{prune} in parallel across the 64 remote R processes. That approach incurs,
in this example, an additional memory overhead of about 16$\,$GB because each
remote R process requires access to a copy of the data matrix $A$. More thrifty
adaptations of the algorithm are possible and could, for example, share the
memory required by $A$ between worker R processes on each computer.

<<echo=FALSE, results=tex>>=
# Default values:
t1=10449.670
t3=482
t4=NA
m1=1261.6
m3=994.3
m4=42282.4

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
 # methylation data
}
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
@
From a set of almost 400 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm also runs, although slower, in
the constrained memory setting of a desktop PC.

%%<<eval=TRUE,echo=FALSE>>=
%%library(irlba)
%%
%%print("R code here")
%%@

@@

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation. Lemma
\ref{sumlemma} decomposes Euclidean distance between two vectors in terms of
singular basis vectors and Algorithm \ref{prune} can be applied directly to the
problem of finding all pairs of vectors that meet a thresholded Euclidean
distance with small changes in steps 3, 4, and 7.  And, by equivalence of
norms, thresholded Euclidean distance can be extended to other distance
metrices similarly. The reference R implementation\cite{sup} includes functions
for general thresholded distance computation.

@@

\bibliography{bibliography}{}
\bibliographystyle{plain}

>>>>

%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installedi in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%% install.packages("xtable",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.

\documentclass[article]{jss}
%% need no \usepackage{Sweave.sty}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}

@@

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
University of Rhode Island
\And
Michael Kane\\
Yale University
\AND
Bryan Lewis\\
Paradigm4, Inc.\\
\And
Alex Poliakov\\
Paradigm4, Inc.
}
\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\Shorttitle{Efficient Thresholded Correlation}

\Abstract{
Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.
}
\Keywords{thresholded correlation, SVD, \proglang{R}}
\Plainkeywords{thresholded correlation, SVD, R}
\Address{
James Baglama, Chair of Mathematics\\
Department of Mathematics\\
University of Rhode Island\\
Kingston, Rhode Island 02881\\
E-mail: \email{jbaglama@math.uri.edu}\\
URL: \url{http://www.math.uri.edu/~jbaglama}\\
}

\begin{document}

@@

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes \citep{genomics}.  Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications \citep{ml1, ml2} and are important to time series and
image processing applications \citep{timeseries, svd-similarity}.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-occurring observations (rows)
among vector pairs; see \citep{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms \citep{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform \citep{timeseries}. The method of Wu, et al. \citep{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

Most of the floating point operations occur in step 4 of Algorithm \ref{prune}.
Fortunately the $\ell$ evaluations in step 4 are independent of each other and
can easily be computed in parallel. See the reference \proglang{R}
implementation \citep{sup} for an example.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
of Baglama and Reichel \citep{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for \proglang{R} \citep{irlbar}, \proglang{Matlab} \citep{irlbam},
and \proglang{Python} \citep{irlbap}.

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without any constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \citep{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

A na\"ive brute force thresholded correlation computation requires $O(n^2 m)$ flops.
Let $p$ be the selected IRLBA dimension. Then Algorithm \ref{prune} requires
$O(n p \ell)$ flops, where $\ell$ is the longest run of ordered entries in
$v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
including the flops required by IRLBA and final thresholding in step 7.  The
main computational savings of $\frac{nm}{lp}$ depends on the rate of decay in
the singular values of $A$, the subspace dimension $p$, and the desired
threshold value. We remark that this estimate of computational savings is
available already in step 3 of Algorithm \ref{prune} after only modest
computational effort. Users might use this cheap estimate to, for example,
alter their threshold in some cases.

@@

We evaluated the algorithm using public gene expression and DNA methylation
datasets from the \proglang{R} {\tt biclust} \citep{biclust} package and from
the Cancer Genome Atlas \citep{gdac}, referred to as TCGA.  Most tests were
performed on a desktop PC with a single AMD A10-7850K 3.7$\,$GHz Athlon
quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running the Ubuntu 15.10
GNU/Linux OS. Tests were performed with 64-bit \proglang{R} version 3.2.3 using
double-precision arithmetic and the OpenBLAS library version 0.2.14-1ubuntu1
(based on Goto's BLAS version 1.13) with {\tt OMP\_NUM\_THREADS=1}. The last
experiment includes a test run on a Linux cluster described in that section.

All examples use the reference \proglang{R} implementation, {\tt tcor()}, which can be
installed from the development GitHub repository using the
{\tt devtools} package with:

@@

The native \proglang{R} {\tt cor()} function is a general function that can compute
several different types of correlation matrices including Pearson, Spearman,
and Kendall and includes a number of options for dealing with missing values.
Because of its flexibility, the native {\tt cor()} function does not always
compute Pearson correlation matrices in the most computationally-efficient way.

@@

\subsection{Small gene expression example}

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as a $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
threshold = 0.95

# Register a four-core multicore parallel backend
registerDoMC(4)

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The following example uses calculated beta values from the Illumina
Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \citep{gdac}.
See the \proglang{R} package implementation \citep{sup} for examples of
efficiently converting the downloaded sample methylation data into an R matrix.
We chose these data for their public availability as a general example of
computing thresholded correlation among sets of large numbers of vectors,
rather than for any specific biological meaning.

@@

Our simple brute force implementation was not possible due to memory constraints.
We remark that the matrix multiplication central to the brute force
implementation can be decomposed into a sequence of smaller steps to
fit within given memory constraints. However, the brute force algorithm would still
need to evaluate all of the nearly 160 billion possible vector pair correlation
values.

@@

We illustrate that steps 4--7 of Algorithm \ref{prune} can be computed in
parallel by comparing the above results run on our reference test PC with
results run on a modest GNU/Linux cluster at Paradigm4, Inc. The cluster
consisted of four computers connected by 10 gigabit ethernet. Each computer
contained two Intel Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per
computer) and 64$\,$GB/1,600$\,$MHz ECC registered RAM running CentOS 6.6 and
64-bit \proglang{R} version 3.2.2 with the reference \proglang{R} BLAS library.

No code change was required to run in parallel on the cluster. The reference \proglang{R}
{\tt tcor()} implementation uses \proglang{R}'s {\tt foreach} \citep{foreach} framework to
run on arbitrary parallel configurations (``back ends''). We simply registered
a {\tt doRedis} \citep{doredis} parallel back end with 16 \proglang{R} workers per computer
(64 total \proglang{R} worker processes) prior to invoking {\tt tcor()}.

The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
below. The cluster example computed the filtering step 7 of Algorithm
\ref{prune} in parallel across the 64 remote \proglang{R} processes. That approach incurs,
in this example, an additional memory overhead of about 16$\,$GB because each
remote \proglang{R} process requires access to a copy of the data matrix $A$. More thrifty
adaptations of the algorithm are possible and could, for example, share the
memory required by $A$ between worker \proglang{R} processes on each computer.

<<echo=FALSE, results=tex>>=
# Test system values:
t1=10309
t3=482
t4=NA
m1=389.1
m3=994.3
m4=42282.4

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  library(doMC)
  system("wget http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/ACC/20150821/gdac.broadinstitute.org_ACC.Merge_methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.Level_3.2015082100.0.0.tar.gz")
  cmd = "cat ACC.methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.data.txt  | tail -n +%s | cut -d '       ' -f 2,6,10,14,18,22,26,30,34,38,42,46,50,54,58,62,66,70,74,78,82,86,90,94,98,102,106,110,114,118,122,126,130,134,138,142,146,150,154,158,162,166,170,174,178,182,186,190,194,198,202,206,210,214,218,222,226,230,234,238,242,246,250,254,258,262,266,270,274,278,282,286,290,294,298,302,306,310,314,318"
  N = 485579  # total number of lines
  h = 2       # total header lines
  # Read the first header line with sample IDs:
  f = pipe(cmd, open="r")
  id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
  close(f)
  # Read the rest of the file in parallel. I prefer doMC/foreach over mclapply for this because foreach can combine results incrementally.
  registerDoMC(4)  # four cores on our test machine
  block = floor((N - h)/cores)
  acc = foreach(j=1:cores, .combine=rbind) %dopar%
  {
    skip = block * (j - 1) + h + 1
    nrows = ifelse(j == cores, -1, block)
    f = pipe(sprintf(cmd, skip), open="r")
    on.exit(close(f))
    read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=nrows)
  }
  acc = t(acc)
  # OK, the example data are parsed and in acc. proceed as in the smaller example.
  threshold = 0.99
  # Compute using tcor():
  m1  = sum(gc()[,2])  # Memory in megabytes currently in use
  t1  = proc.time()
  tx  = tcor(acc, t=threshold, p=10)
  t1  = (proc.time() - t1)[3] # Time in seconds
  m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test

  # The cluster example proceeds similarly, but registers a doRedis backend
  # instead of the doMC backend before invoking tcor...
} # end if run examples
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
@
From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

@@

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation. Lemma
\ref{sumlemma} decomposes Euclidean distance between two vectors in terms of
singular basis vectors and Algorithm \ref{prune} can be applied directly to the
problem of finding all pairs of vectors that meet a thresholded Euclidean
distance with small changes in steps 3, 4, and 7.  By equivalence of norms,
thresholded Euclidean distance can be extended to other distance metrics. The
reference \proglang{R} implementation \citep{sup} includes functions for thresholded
distance and correlation.

The ideas presented here can be adapted to efficiently find the top N most
correlated vector pairs for a given constant N similarly to the work of
Xiong et al. \citep{prune2}.

@@

We present a simple algorithm and reference \proglang{R} implementation based on a
truncated singular value decomposition (SVD) that efficiently computes
thresholded subsets of large correlation and distance matrices.  We show that
the natural connection between data covariance and the SVD can result in
substantial computational and memory savings, and illustrate these savings with
a few genomics data experiments.  Moreover, the main computational work in the
algorithm is easily parallelized and easily deployed across computational
clusters.

\bibliography{bibliography}{}
